|

### 2021-09-05 article/article.Rmd
changes paper

---
title:  |
        | Legislature Size and Welfare:
        | Evidence from Brazil
author: 
  - Gabriel Cepaluni[^gabriel]
  - Danilo Freire[^danilo]
  - Umberto Mignozzetti[^umberto]
date: \today
thanks: "We thank Cedric Antunes, Victor Araujo, Eric Arias, Taylor Boas, Natalia Bueno, Andrea Freitas, Saad Gulzar, James Mahon, Lucas Novaes, Nelson Ruiz, Catarina Roman, David Skarbek, and Denis Stukal for their valuable comments. We also thank the participants of seminars and conferences at CEPESP, CUNY, INSPER, IPSA, LACEA, LASA, MPSA, Notre Dame, NYU, PUC Chile, SPSA, and Uppsala. We gratefully acknowledge the financial support of the São Paulo Research Foundation (FAPESP), grant number 2017/07290-5. All materials required to replicate this study are available at <https://github.com/umbertomig/legislature-size-welfare>."
abstract: "What is the effect of legislature size on social welfare? While the literature suggests that increasing the number of legislators broadens political representation and raises government expenditures, its implications for public service provision remain understudied. In this paper, we exploit exogenous changes in city-council size in Brazil and show that adding a legislator increases education and healthcare indicators. XXXX Education quality and preventive health care remain unaffected, but primary school enrolment and infant mortality significantly improve. To investigate the mechanism, we surveyed former councillors and analysed 346,553 bills proposed between 2005 and 2008. This analysis shows that politicians prefer to provide private and targeted goods instead of public goods. This paper has implications for the design of legislative institutions."
abstractspacing: double
keywords: Legislature size; Legislative politics; Public goods provision; Regression discontinuity design; Brazil
jelcodes: HXXXX
fontsize: 11pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
# XXXX
```

@@

Legislative institutions are crucial for welfare and service provision in democratic countries. Historically, the division of power and the birth of a representative government placed the legislatures at the center of popular consent \citep{north1989constitutions}. In modern democracies, most countries have assemblies at all levels of government. These institutions affect the information and quality of legislation \citep{krehbiel2010information}, check and veto the other political powers \citep{tsebelis2002veto}, and influence policymaking \citep{weingast1988industrial, north1989constitutions, auriol2012optimal}. Moreover, they also determine taxation, government expenditure \citep{Weingastetal1981, PrimoSnyder2008, stasavage2011states}, and oversee service provision \citep{mccubbins1984congressional, poulsen2018political}.

Regardless of differences in electoral rules, social composition, and procedures, all legislatures share one feature: more than one representative composes them. Legislature size relates to collective action problems \citep{crain1979cost, rogers2002free}, government spending \citep{Weingastetal1981, PrimoSnyder2008}, and representation \citep{allenrepresentation}, but how these components affect welfare remain understudied. On the one hand, enlarged legislatures increase government spending and lawmaking, and this may improve some services \citep{Weingastetal1981, }. On the other hand, adding a legislator can increase the number of veto players and broaden the collective action problems, hindering service provision. Therefore, larger legislatures have an ambiguous effect on welfare, and XXXX.

In this paper, we propose a theory of legislature size and welfare that takes into account two crucial components: the XXXX and the partisan affiliation of the legislators.

XXXX Umberto: Parei aqui.

This paper fills this gap by studying the effects of legislature size on service provision. Assuming that politicians want to further their careers, we argue that two variables determine service provision: the perceived importance of the service by voters and the politician's capacity to claim credit for its provision. Politicians perform a simple cost-benefit analysis: if providing a service results in high electoral yield, and it is easy to claim credit for, politicians will produce this service abundantly. When a service is hard to claim credit for, politicians only invest their time if it is highly preferred by voters. Finally, when services are both hard to claim credit for and of low importance to voters, they are placed at the bottom of the politicians' priorities.

To test this theory, we exploit an exogenous variation in Brazilian city council sizes. In 2004, the judiciary reinterpreted the constitution, unexpectedly changing the number of city councilors in Brazilian municipalities. Before the decision, towns could freely choose their legislature size, but in March 2004, the Superior Electoral Court created population thresholds, adding one councilor for each set of 47,619 inhabitants. Around the cutoffs, this decision represents an exogenous change in city council size, allowing us to investigate the welfare effects of increasing the legislatures in Brazil.

Studying the 2005 to 2008 Brazilian mayoral term, when the ruling was in effect, we find improvements in health care and education. Increasing the council size by one legislator lowers infant mortality by 1.99 per one thousand infants born and increases elementary school enrollment by 2.58 children in kindergarten classrooms. However, the additional legislator had no influence on preventive health care and education quality. [] Voters place a high electoral premium on hospital care and mediating school enrollment is relatively easy to claim credit for. In line with our theoretical model, these services should improve with legislative size. On the other hand, preventive health care is harder to claim credit for, and school quality improvements are expensive and pay off only in the long-run. Thus, these services remain unchanged with larger legislatures.

Finally, we present three sets of evidence that the results are consistent with the micro-level representation practices of city councilors. First, we show that the new councilor has access to government resources and uses public office appointments to mediate service provision. They have a 91\% chance of belonging to the mayoral coalition, and they appoint an average of 105 extra bureaucrats for many positions at the mayor's office, having loyal workers inside the municipal bureaucracy. Second, we analyzed 346,553 bills approved by city councilors in municipalities close to the council size thresholds. In line with our theory, we find that councilors prefer to provide targeted goods (72.9\% of all bills), as they satisfy the populations' needs and are easier to credit claim. Finally, we ran an online survey among the councilors during the period, showing that most councilors believe that voters prefer targeted services (68.8\%) to lawmaking (24.6\%) and oversight (30\%).\footnote{Councilors were able to select multiple categories, so the sum can be over 100\%.} We also refute three alternative explanations: increased representation of women and non-white legislators, change in electoral competitiveness, and a larger average of approved legislation by each councilor.

# Legislature Size, Legislative Support, and Welfare
\label{sec:theory}

What is the effect of larger legislatures on public service provision? The most striking feature of a legislature is the fact that decisions are undertaken collectively. Legislators have to propose changes in the status-quo, agree upon what they want to decide, evaluate proposals put forward by mayors, discuss and improve the proposals, and decide which changes they will carry on.

In this sense, legislature size is critical in determining the efficiency of the decision-making process. Upon adding a new city councilor, the most simple change is the increased production capability. Legislative production, analogous to what happens in firms, may increase when adding an extra legislator \citep{crain1979cost, crain1982team, rogers2002free}. For instance, suppose that working full time, politicians can approve five bills a month. Ten politicians can together approve 50 bills, while 11 can pass 55. This increases the number of measures adopted, and if laws enact services to constituencies, then we shall witness an increase in service provision.\footnote{An alternative explanation relates to the Condorcet Jury Theorem. In these models, there is a binary choice, with one of them being ex-ante the best. If legislators access the best choice with a probability higher than 0.5, more legislators will increase the change of selecting the best choice \citep{myerson1998extended}. However, this logic assumes that politicians choose among exogenously provided proposals, which makes sense in the case of a jury, but is somewhat uncommon for legislatures. \citet{auriol2012optimal} compute the optimal size of chambers considering that this is the solution to the interplay between specialization costs and citizens' preferences.}

However, increasing the size of legislatures also carries considerable costs. Larger legislatures can raise the transaction costs \citep{weingast1988industrial}, increase collective action problems \citep{crain1979cost}, and make  forming majorities more costly \citep{crain1982team}. These negative features, together with the fact that legislators may free-ride on each others' proposals, may decrease legislative productivity. For instance, if decisions need a majority or qualified majority to be approved, more legislators create a perverse incentive, increasing the difficulty of approval. Moreover, legislators are pressured to provide targeted services for their constituencies at the expense of the entire polity \citep{Weingastetal1981, PrimoSnyder2008}, and this can raise inefficiencies in expenditure allocation. Additionally, transaction costs within legislatures may increase the difficulty of passing bills \citep{weingast1988industrial, baron1989bargaining}. Therefore, it is unclear whether and how larger legislatures impact lawmaking and service provision.

In our view, to understand the effects of larger legislatures, we need to consider the political motivations behind the legislator's work. While an extra politician may increase the productivity frontier of a legislature, the types of services that will be favored remain unclear. To understand the political motivations, we need a theory that accounts for the cost-benefit calculations performed by politicians.

When deciding which service to prioritize, politicians look into two dimensions of the service provision. First, the competition with other legislators motivates politicians to provide services that they believe voters want, regardless of whether these policies improve long-term welfare. For instance, suppose that there are three services, organized hypothetically based on beliefs about which of these three voters are more likely to reward:\footnote{Note that there is a difference between \textit{what voters want} and what \textit{politicians believe voters want}. Although politicians and brokers mostly access voter's beliefs accurately \citep{FinanSchechter2012, Stokesetal2013}, the crucial signal that a politician gets is vote shares, which are aggregated. It is hard to disentangle which service generated the highest electoral yield for the politician. Therefore, the politician's beliefs play an essential role in the services that they prioritize.} improving a local health clinic (most preferred), building a playground (second most preferred), or increasing the quality of education (least preferable). Politicians seeking recognition will mostly want to improve the health care system, which they believe will most benefit their electoral yields. After getting this done, they will shift efforts toward the playground construction and finally to education quality. Therefore, increasing legislature size will disproportionately increase the provision of services that the politicians believe voters prefer and will reward the most in the polls.

Second, to get reelected, besides providing the services that voters prefer, politicians have to take credit for facilitating provision \citep{fiorina1987personal, gulzar2017politicians, nielsen2017politicians, silva2017clarity}. Typically, the services that are easiest in terms of credit claiming are either providing local public goods, such as improving a neighborhood health clinic, or purely clientelistic and personalistic services \citep{Weingastetal1981, kuschnir2000cotidiano, Stokes2005, nichter2011electoral, Stokesetal2013, luna2014segmented, vieira2015clientelismo, bertholini2018pork}. This explains why clientelism and pork-barrel are persistent in democratic polities. Table \ref{tabcli} summarizes the incentives.

\begin{table}[htp!]
\centering
\caption{Effects of Legislature Size on Service Provision}
\label{tabcli}
\scalebox{1}{%
\begin{tabular}{@{}lcc@{}}
\toprule
& \textbf{Hard for Credit Claim} & \textbf{Easy for Credit Claim} \\ 
\midrule
\multirow{3}{*}{\textbf{Low ranked by voters}} & No changes & None or mild \\
& or deterioration & improvements \\
& (Preventive Health)  & (Honors bills) \\[0.5em]
\multirow{3}{*}{\textbf{High ranked by Voters}} & Mild to low & Mild to strong \\ 
& improvement & improvement \\
& (Infant Mortality)  & (School Enrollment) \\
\bottomrule
\end{tabular}
}
\end{table}

These two dimensions provide a simple yet powerful guide of the politicians' incentives. When voter places a high premium on the provision of a given service, and it is easy to claim credit for the provision, politicians will concentrate their efforts in providing it. The more politicians, the higher the provision of the service. Consider, for instance, constituency service, such as getting a child enrolled at a public school. The voters highly rank service such as this, and the politicians can break the school access directly, getting all the credit for the provision. Therefore, these services tend to increase, up to the point of being over provided.

When a service is hard to credit claim, to be provided, it has to be highly ranked by the voters. Consider services that decrease infant mortality. Although policies in this direction might be harder to be directly associated with one given politician, every politician wants to be known as the ones that diminished infant mortality. The death of a child has devastating effects over a polity, and no politician wants to be associated with it. On the other hand, every parent would be extremely thankful for the politicians that improved health care and saved their child. Moreover, politicians might facilitate access to medication and hospitals, which in turn is easier for credit claiming. Therefore, these services will have anywhere from modest to substantial improvements.

Suppose now that service is secure to credit claim for, but gives low electoral yield. In this case, although the provision can be unquestionably traced back to the legislator, voters care little about the provision. For instance, honors and appraisal legislation are very common in Brazilian municipalities. However, they are perceived by politicians as low electoral yield, but the legislator that proposed the homage usually gets associated with it. These types of legislation increase with legislature size, but their effect on welfare is negligible.

Finally, services that are difficult to credit-claim, or that politicians believe that voters care very little about, are expected to receive little invested effort from politicians. For instance, consider preventive health care. Preventive care refers to health measures, such as prenatal care or vaccination, undertaken to lower the chances or the severity of a disease(s) in the future. However, it is a compounded chance: the voter has to consume the service, but its effectiveness depends on voters getting sick. Nowadays, we are witnessing how hard it is to vaccinate children, despite all the scientific evidence demonstrating its effectiveness. Moreover, rational voters may fail to punish politicians for low investments in preventive public policies, such as natural disasters or climate change mitigation \citep{gailmard2019preventing}. Hence, voters might also not punish politicians for low investments in preventive health care because they might only observe whether a politician properly invested in preventive public policies in cases of emergency. Therefore, we shall expect little to no improvement in these types of services with larger legislatures.

In sum, these considerations suggest that instead of looking for overall positive or negative effects, the effects of legislature size should manifest diversely, conditional on the type of service studied. The most significant changes should be expected in services that voters want and that are easier to claim credit for. Services with only one of these characteristics will be intermediate in terms of their priority and efforts exerted by politicians in their provision and will improve with less intensity. Services hard to credit claim and low ranked by voters will tend to be unaffected or under-provided by lawmakers.

>>>>

---
title:  |
        | Legislature Size and Welfare:
        | Evidence from Brazil
author: 
  - Gabriel Cepaluni[^gabriel]
  - Danilo Freire[^danilo]
  - Umberto Mignozzetti[^umberto]
date: \today
thanks: "We thank Cedric Antunes, Victor Araujo, Eric Arias, Taylor Boas, Natalia Bueno, Andrea Freitas, Saad Gulzar, James Mahon, Lucas Novaes, Nelson Ruiz, Catarina Roman, David Skarbek, and Denis Stukal for their valuable comments. We also thank the participants of seminars and conferences at CEPESP, CUNY, INSPER, IPSA, LACEA, LASA, MPSA, Notre Dame, NYU, PUC Chile, SPSA, and Uppsala. We gratefully acknowledge the financial support of the São Paulo Research Foundation (FAPESP), grant number 2017/07290-5. All materials required to replicate this study are available at <https://github.com/umbertomig/legislature-size-welfare>."
abstract: "What is the effect of the number of legislators on public service provision? Despite the importance of institution designs for young democracies, little is known about the particular features that may help improve welfare. This paper uses a formal model to show how increasing legislature size may affect service provision. I show that changes in bargaining costs depend on whether the additional legislator shares the Executive's party affiliation: More opposition members reduce the Executive's equilibrium public goods provision, while more government-aligned members increase it. I test this theory by exploiting sharp discontinuities in city council size in Brazil. I show that an additional city councilor has a 91% chance of belonging to the mayoral coalition, significantly improving primary school enrollment and infant mortality. To investigate the mechanism, I surveyed former councilors and analyzed 346,553 bills proposed between 2005 and 2008. This paper has implications for the design of representative institutions."
abstractspacing: double
keywords: Legislature Size; Legislative Bargaining; Local Public Goods Provision; Regression Discontinuity Design; Brazil
jelcodes: H41; H80; I31; I38
fontsize: 11pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

@@

Legislative institutions are crucial for welfare and service provision in democratic countries. Historically, the division of power and the birth of a representative government placed the legislatures at the center of popular consent. In modern democracies, most countries have assemblies at all levels of government. These institutions affect the information and quality of legislation, check and veto the other political powers, and influence policymaking. Moreover, they also determine taxation, government spending and oversee service provision \citep{Weingastetal1981, weingast1988industrial, north1989constitutions, auriol2012optimal, stasavage2011states, poulsen2018political}.

Regardless of differences in electoral rules, social composition, and procedures, all legislatures share one feature: more than one representative composes them. Legislature size relates to collective action problems, bargaining between political powers, government spending, and representation, but how these components affect welfare remain understudied \citep{crain1979cost, Weingastetal1981, rogers2002free, PrimoSnyder2008, allenrepresentation}. On the one hand, enlarged legislatures may increase government spending and lawmaking, and this could improve some services \citep{Weingastetal1981}. On the other hand, adding a legislator may increase the number of veto players and broaden the collective action problems, hindering service provision \citep{tsebelis2002veto, gottlieb2019countervailing}. Therefore, what is the effect of increasing the legislature size on public service provision?

In this paper, I propose a local legislature size and welfare theory that incorporates two crucial features: Executive leadership and party politics. First, in mayor-council municipalities, mayors tend to be central in the local public service decisions. Mayors have the prerogative in proposing public investments, much as presidents in presidential systems. Second, partisanship is essential at the local level, even in settings where the ideological attachment is weak, such as in most young democracies. Party attachment helps structures political views \citep{samuels2014power}, and local-level elites usually belong to different factions and parties. They fight for power in the municipality by controlling various parties and forming distinct pre-electoral coalitions.\footnote{Elite political competition in the local level has also been linked to violence and conflict in Brazil and elsewhere \citep{hoelscher2015politics, mignozzetti2018predatory}.} In most cities, family ties are substitutes for the ideological component, but the most influential local leaders sort themselves in different political groups \citep{nunes1977coronelismo, ferraz2020political}.

This paper incorporates partisanship and mayoral first-move advantage to a \citet{baron1989bargaining} bargaining model. Mayors make the initial policy proposal that consists of an investment in public goods provision, a decision to extract rents for personal gains, and a vector of side-transfers for councilors. If they fail to secure majoritarian support from councilors, a reversal stage occurs, where the decision leaves the mayor's office, and councilors then propose the public investments. In this mayor-council bargaining, costs come from the side-transfers required to secure a majority among councilors. I show that in a model without partisan concerns, bargaining costs always increase with larger city council size, lowering public goods provision and rents. However, adding partisanship decreases the bargaining costs when the chances of electing councilors aligned with the mayor are sufficiently high. Lower bargaining costs release funds for mayors to invest in public goods provision. In this case, a larger city council may have more partisan councilors aligned with the mayor, lowering the mayor's bargaining costs and increasing the equilibrium public goods provision.

I test these theoretical predictions by exploiting an exogenous variation in Brazilian city council sizes. In 2004, the judiciary reinterpreted the Constitution, unexpectedly changing the number of city councilors. Before the decision, towns could freely choose their legislature size, but in March 2004, the Superior Electoral Court created population thresholds, adding one councilor for each set of 47,619 inhabitants. Around these cutoffs, the decision represented an exogenous change in city council size, allowing me to investigate the effects of increasing the number of local legislators in Brazil.

Studying 2005 to 2008 Brazilian mayoral term, I find that increasing city council size changed the composition of the mayoral coalition and the number of appointed bureaucrats in the municipality. The extra city councilor had a 91 percent chance of belonging to the mayor's pre-electoral coalition. This enhanced chances of electing a councilor connected with the mayor decreased the bargaining costs for the mayor, freeing investments for public services. Moreover, the number of appointed bureaucrats increased by 104 with the additional councilor. In Brazil, more appointed bureaucrats lead to higher quality workers, better service provision, and more accountability \citep{toral2019benefits}.

More support for the mayors and more politically appointed bureaucrats correlate with city councilors' service provision expectations and legislation approved in Brazilian municipalities. I surveyed 174 former councilors who served during the 2005-2008 term, and they corroborate that mayors use bureaucratic appointments and granting favors to secure support. As the formal model shows, bargaining with councilors has a resource transfer component. By investigating what councilors prefer to provide to voters, I find that councilors demand job appointments, as these are helpful for their reelection chances and improve municipal welfare. These findings align with a novel dataset I build with 346,553 laws passed in 63 municipalities within ten thousand inhabitants away from the cutoffs. The legislation in municipalities on the left and on the right-hand side of the cutoffs comprised local public goods. However, municipalities with larger city councils had around 15.0 percent more public goods provision proposals, suggesting an increase in service provision legislation.

Finally, I show that the increase in legislature size significantly impacts municipal education and health care. Increasing the city council size by one legislator lowers infant mortality by 2.01 children per one thousand infants born and reduces postneonatal mortality by 0.90 children per one thousand infants that survived their first 28 days. A larger council size also increased enrollment by 2.58 children in elementary education classrooms without significantly diminish elementary school quality. These services are crucial for welfare and result from an eased bargaining between the mayor and the city council.

This paper contributes to three strands of the literature. First, this paper contributes to a large literature on multilateral bargaining. Repeated models of multilateral bargaining started with \citet{baron1989bargaining}, which extended the bilateral bargaining game proposed by \citet{rubinstein1982perfect} to more than two players. Several papers followed \citet{baron1989bargaining}, adding endogenous recognition power, changing the size of the required majority, adding partisanship, and testing the predictions of these models in the lab \citep{calvert2005legislative, choate2019partisan, agranov2020persistence, baranski2020determinants}.\footnote{See \citet{eraslan2019legislative} for a comprehensive review of the multilateral bargaining and \citet{ray2015coalition} for a review on coalition formation.} In this paper, I extend the \citet{baron1989bargaining} model under closed rules, adding the mayor's first move and political partisanship. Partisanship here is different from \citet{choate2019partisan} since I consider partisan as a zero-sum effect that is positive for politicians aligned with the mayor and negative otherwise. This paper investigates the often overlooked effect of legislature size on the bargaining equilibrium strategies.

Second, this paper contributes to the literature regarding the political and economic effects of larger legislatures. Following the contribution by \citet{Weingastetal1981}, the "law of 1/n" posits that larger legislatures increase the legislative pressure for providing targeted public goods to local supporters at the expense of the entire polity \citep{alptekin2021effect}. The corollary of the "law of 1/n" is that increasing legislature size should promote excessive public spending.\footnote{\citet{PrimoSnyder2008} show that legislature size may result in positive or negative expenditures, depending on factors such as the degree that local public goods can exclude non-locals or deadweight costs associated with taxation.} In the US metaphor, this would explain significant infrastructure spending in smaller districts. However, the "law of 1/n" literature left aside the public goods provision and welfare. This paper is one of the first papers to study the number of local legislators on welfare. The model provides a theoretical expectation based on the chances of electing a politician aligned with the mayor. The empirical estimates confirm the theoretical predictions, with the additional fact that legislature size has a null effect on expenditures in Brazil.

Finally, it contributes to a growing literature on municipal partisan politics. Most papers in this area estimate the effects of the mayor's party on political outcomes. For instance, \citet{gerber2011mayors} and \citet{de2016mayoral} find differences in governance when comparing Democratic and Republican mayors. Specifically about Brazil, \citet{gouvea2021partisanship} shows that social policy expenditure increase with left-wing mayors. \citet{frey2020larger}, using the discontinuities in city council size after the 2009 constitutional amendment that reverted the thresholds I study here, shows that changes in council size lower the support for candidates backed by the municipal mayors. This paper reinforces the view that partisanship matters locally, even in countries with weak ideological linkage.

# A Theory of Legislature Size, Legislative Bargaining, and Welfare
\label{sec:theory}

Several cities in the world are governed by the mayor-council system. In mayor-council cities, the power is divided between a mayor that controls the Executive branch and a city council that supports or sanctions the policies proposed by the mayor. In this setting, consider an strategic interaction between a mayor ($M$) and $N \geq 3$ city councilors, indexed in $i \in \{1, 2, \cdots, N \}$. The mayor has the prerogative of proposing a vector of policies that is voted by the city councilors. The provision of public goods is then a bargain between the mayor and the city council. Policy proposals are a combination of public goods provision $g$, rents for the mayor $r$, and a vector of transfer for city councilors \textbf{x}.

The city council votes the mayor's proposal, and if councilors accept it, the mayor implements it. Otherwise, a reversal policy takes place. Reversal policy in this context means that the mayor's proposal failed to secure the majority of council votes. The decision process leaves the mayor's office, and it is transferred to the council. The types of reversal policies are crucial for my argument, as it changes the relative strength of the council in their bargaining with the mayor. I investigate two types of reversal policies: a baseline reversal with no parties and a hybrid mechanism with partisan and non-partisan concerns. The differences between these two mechanisms are that the non-partisan councilors only care about rents, while partisan councilors care about both rents and policy.

To govern, the mayor must convince at least half of the city councilors to support her policy proposals. Similar to \citet{buenodemesquita2005logic}, I assume that the selectorate, i. e., the individuals that have a say in the mayor's policy proposals, are the $N$ city councilors. The winning coalition, the minimum number of members that must be convinced to implement the policy, is equal to half of the selectorate $N/2$. City councilors are moved by policy concerns $p$ and transfers received from the mayor or the councilor recognized as the proposer \textbf{x}. I leave the definition of these transfers open, as this facilitates the relationship with several instruments that mayors can use to pressure city councilors. Transfers may be granting electoral favors, favoring clientelistic brokers, spending in areas that the councilor has political or personal interests, providing portfolio (municipal offices) and public jobs within the municipal bureaucracy, or even campaign contributions and bribes. The use of transfers to affect the councilors' choices generate a bargaining cost of $C_G$. The bargaining costs vary with the reversal mechanism and the legislature size.

Finally, after learning the bargaining costs, the mayor decides a level of public goods provision ($g$) and pockets rents ($r$). Public goods provision helps the mayor to get reelected. Rents are for direct mayor's consumption, and they do not contribute to the mayor's electoral success.\footnote{This model, similarly to \citet{cowen1994rent} and \citet{buenodemesquita2005logic} combines public goods provision and rent extraction. For a dynamic bargaining model that incorporates transfers to districts and public goods provision, see \citet{battaglini2012legislative}.} This makes the mayor's expected utility a sum of the gains from rents and the benefits from reelection. Both the utilities from rents $u(.)$ and the probability of reelection $\pi(.)$ are concave functions, meaning that more rents or public goods increase the utility at a decreasing rate. The probability of reelection is multiplied by the benefit from holding office $B_M > 0$. This benefit captures the tangible and intangible gains that the mayor perceives from holding the public office. I assume that the benefits are high enough to rule out an equilibrium with only rents and transfers, which is equivalent to saying that mayors strongly prefer to be reelected. If the mayor's policy proposal is approved, then the expected utility for the mayor is:

\[
\mathbb{E}U(r, g) \ = \ u(r) + B_M\pi(g)
\]

The policy choices of the mayor are subject to the municipal budget constraint. The municipality has $R > 0$ resources, and cannot run debts, meaning that the budget must be balanced. I also assume that there are enough resources for the mayor to govern.

Let $C_G$ be the bargaining costs. The budget balance constraint requires the offers that the mayor makes to the city councilors to satisfy the following inequality:

\[
r + g + C_G \ \leq \ R
\]

The expected utility for the city councilors depends on the reversal policy. In the baseline non-partisan reversal, when the council rejects the mayor's proposal, a reversal stage starts with resources diminishing by a factor of $\delta \in (0,1)$ and the selection of one councilor randomly. The selected councilor becomes the new proposer. If her proposal is accepted, it is implemented. If councilors reject the offer again, the budget shrinks by a factor $\delta$, and another city councilor is recognized to propose. The process repeats until a proposal is finally accepted.

After studying the non-partisan baseline, I consider a situation that combines both partisan and non-partisan legislature concerns. Each councilor has a party affiliation. Party affiliations are mutually exclusive, and if a city councilor is aligned with the mayor's party, then she belongs to the government coalition $G \subset \{1, 2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O = \{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political value of $p > 0$ for the councilors, and whether it adds or subtracts from the councilor's utility depends on the party affiliation of the councilor. If councilors reject the mayor's proposal, no public goods are provided ($p=0$). Moreover, in equilibrium, I assume that city councilors would prefer to receive transfers instead of providing public goods.\footnote{In the online supplemental materials, I consider a version of the model where public goods provision depends on the party affiliation of the proposer so that it affect choices in any potential history of the game. The equilibrium public goods provision level is even higher than in the model presented here, as city councilors prefer to approve the mayor's proposal than run the risk of an opposition member becoming the proposer.}

The timeline of the strategic interaction is as follows:

\begin{enumerate}
  \item The mayor learns how many government $|G|$ and opposition $|O|$ legislators were elected.
  \item The mayor proposes a policy vector $(r, g, x)_M$.
  \item The city council votes the proposal.
  \begin{itemize}
    \item If the council accepts, the policy is implemented, and the game ends. 
    \item Otherwise, the reversal policy is implemented.
  \end{itemize}
  \item (Reversal Policy) The budget is discounted by a factor $\delta \in (0,1)$, and one councilor is randomly selected to make an offer. If the offer is accepted, it is implemented. If rejected, the reversal policy stage restart.\footnote{In section B.1 of the online supplemental materials, I solved three reversal mechanisms. The two in here, plus a mechanism with only political concerns.}
\end{enumerate}

The solution concept I use in this model is the Subgame Perfect Nash Equilibrium. A Subgame Perfect Nash Equilibrium requires that the strategies played are a Nash Equilibrium in each of the subgames of the game. The solution is found using backward induction. In the model with infinitely repeated proposals, any strategy can be shown to be a Nash equilibrium. Thus, a natural assumption is to extend the equilibrium concept to require the equilibrium to be stationary. A Stationary Subgame Perfect Equilibrium requires that, at each given point of the game, if a politician accepts an offer at time $k+1$, she should take the same offer at time $k$ \citep{rubinstein1982perfect}. Stationarity gives us an objective of finding a proposal that would be accepted at any stage in the game. Applying this reasoning, I can characterize a sequence of offers for each point $k$ in time. For the mayor, the optimal offer is at $k=0$, representing a no delay in policy implementation. Moreover, I assume that the players play no dominated strategies, even when there is less than sufficient support for implementing the policy.

For the solution to be of general interest, I solve bargaining costs for the expected number of opposition and government-aligned city councilors. This is because the solution for each fixed number of opposition and government councilors is trivial and does not help understand the overall relationship between council size and bargaining costs.

## The Mayor's Decision Stage

Solving the game by backward induction requires us to start with the councilor's strategies. However, for exposure convenience, assume that I solved the game for the councilors and found the equilibrium bargaining costs $C_G(N)$ when there are $N$ city councilors. I now derive the optimal rents ($r$) and public goods provision ($g$) proposed by the mayor. The mayor benefits from public goods provision, as it increases the chances of her reelection. However, she prefers to invest as minimum as necessary in public goods and extract the remaining resources as political rents. The mayor's objective is to maximize her expected utility, subject to the municipal budget constraint.

\[
\begin{aligned}
\max_{r, g} \quad & u(r) + B_M\pi(g)\\
\textrm{s.t.} \quad & r + g + C_G(N) \leq R
\end{aligned}
\]

In equilibrium, the budget binds, and the first-order condition for an optimal public goods provision makes the marginal benefits to increase rents, in this case, marginal costs of providing the public goods, equal to the marginal benefits of reelection:

\[
u'(R - g - C_G(N)) = B_M\pi'(g)
\]

The optimal level of public goods provision equates the marginal gains accrued from rents with the marginal benefits from reelection.

\begin{proposition}\label{prop1}
The provision of public goods increases with legislature size when the bargaining costs decrease with legislature size.
\end{proposition}

All proofs are in section B.1 of the online supplemental materials. The proof's intuition involves finding what happens to the mayor's marginal utility when I increase both $g$ and $N$. As $N$ increases discretely on a complete lattice, I use monotone comparative statics to derive the conditions for increasing differences \citep{milgrom1994monotone, ashworth2006monotone}. Increasing differences mean that higher $N$ leads to higher equilibrium $g$.

Proposition \ref{prop1} provides the first empirical hypothesis to test: \textit{if the bargaining costs decrease when the legislature size increases, then the provision of the public goods increases when the legislature size increase}. Now I study each reversal mechanism to determine the condition under which the bargaining costs decrease when legislature size increases.

## Baseline: Non-Partisan Legislative Bargaining

In the non-partisan reversal, I assume that when the council rejects the mayor's office, a reversal stage starts with the selection of one councilor randomly. This formulation is similar to the closed rules proposal in \citet{baron1989bargaining}. To find a Stationary Subgame Perfect Nash equilibrium, suppose that there were $k-1$ rejections, and the game is at the $k$-th stage. A councilor accepts the proposer's offer if, and only if, to accept the offer at $k$ is better than or equal to wait until the next stage $k+1$. If the offer is $x_i$, then:

\[
x_i \geq \dfrac{1}{N}\left[ \delta^{k+1}R - \dfrac{N}{2}x_i \right] + \left[ 1 - \dfrac{1}{N} \right]\left[\dfrac{1}{2} x_i \right]
\]

On the left-hand side, I place the offer. On the right-hand side, there are two components. The first is the amount that the councilor $i$ gets when she is the proposer. It is equal to the chance that she is recognized as the proposer times the budget in the next round minus the offers she makes to convince half of the councilors. The second part is the gains if the councilor rejects the current received offer but still gets an offer in the following round. It is equal to the chance that she is not recognized as the proposer times the chance that she receives a transfer times the transfer amount. Note that the offer the councilor makes as the proposer is the same as the offer she wants to receive. This is because the city councilors are exchangeable. The solution is symmetric for all councilors receiving an offer (this means that, without loss of generality, I could have dropped the $i$ in the solution). After some algebra, the offer $x_i$ must be greater than or equal to:

\[
x_i \geq \dfrac{2 \delta^{k+1}R}{2N + 1} \equiv \underline{x}(k, N)
\]

The proposer always offers the minimum required to get the proposal approved. In this case, the offer at any given stage $k$ is going to be equal to $\underline{x}(k, N)$.

Proceeding backwards, at the mayor's proposal stage $k=0$ the mayor is going to offer $\underline{x}(0, N) = \dfrac{2 \delta R}{2N + 1}$ to half of the councilors. In this context, the bargaining costs for the mayor are:

\[
C_G(N) \ = \ \dfrac{N}{2}\left[\dfrac{2 \delta R }{2N + 1}\right] \ = \ \dfrac{\delta R N}{2N + 1}
\]

\begin{proposition}
In the baseline non-partisan reversal, the bargaining costs always increase when the size of the legislature increases.
\end{proposition}

The intuition for the results is that every additional councilor will demand more resources from the mayor. Thus, more councilors represent higher bargaining costs. And if the bargaining costs increase with legislature size, then any increase in public goods provision associated with legislature size has come from a different mechanism. Below I show that partisanship is the key to understand how larger legislature sizes can increase public goods provision.

## Partisan and Non-Partisan Bargaining

In a reversal mechanism that incorporates partisan and non-partisan motivations, a city councilor aligned with the government favors the mayor's offer if:

\[
x_i \geq \dfrac{2\delta R}{2N+1} - p
\]

An opposition politician, on the other hand, favors the mayor's offer if, and only if:

\[
x_i \geq \dfrac{2\delta R}{2N+1} + p
\]

I decompose these costs into two components—first, the costs in terms of rents. Second, the costs (or benefits) from political alignment. In this context, the bargaining costs depend on the mayor's offers for councilors in the mayor's coalition versus the mayor's proposals for councilors in the opposition. The ex-ante chance of an opposition member be elected is equal to $1-\gamma$, and taking the weighted averages, these costs become:

\[
C_G(N) \ = \ \dfrac{N}{2}\left( \dfrac{2\delta R}{2N+1} - p \right) + \left(\dfrac{N}{2} - \gamma N \right)2p
\]

\begin{proposition}
In the reversal mechanism that incorporates partisanship, if $\gamma \geq \dfrac{1}{p}\left[ \dfrac{1}{(2N+1)(2N+3)} \right] \equiv \underline{\gamma}$, then the bargaining costs decrease when the legislature size increases.
\end{proposition}
